---
output: rmarkdown::github_document
---

# tsgp

Simple, lightweight R package for fitting, visualising, and predicting time-series data using Gaussian processes.

```{r, include = FALSE}
knitr::opts_chunk$set(
comment = NA, fig.width = 7, fig.height = 5, cache = FALSE)
```

## Installation

You can install the development version of `tsgp` from GitHub using the following:

```{r eval = FALSE}
devtools::install_github("hendersontrent/tsgp")
```

## Premise

`tsgp` implements the functionality presented in a [recent tutorial](https://hendersontrent.github.io/posts/2024/05/gaussian-process-time-series/) for modelling time-series data with Gaussian processes. Currently, only the univariate setting is supported. `tsgp` works on a [structural time series](https://www.sciencedirect.com/science/article/abs/pii/S0169716105800458) perspective. That is, by decomposing a time series into its constituent statistical parts (e.g., trend, seasonality, noise), one can model each component independently before combining them to form the complete picture of temporal dynamics. This is not only intuitive, but it is also highly transparent---meaning that intelligent and justifiable modelling decisions must be made in order to appropriately capture the data generating process.

`tsgp` is extremely lightweight in both its dependencies and computational approach. If you are seeking a more rigorous or flexible approach to using GPs for time-series analysis, please look into [`Stan`](https://mc-stan.org), [`GPy`](https://gpy.readthedocs.io/en/deploy/), [`GauPro`](https://github.com/CollinErickson/GauPro), [Tensorflow Probability](https://www.tensorflow.org/probability), or [`GaussianProcesses.jl`](https://github.com/STOR-i/GaussianProcesses.jl).

## Functionality

Currently, `tsgp` supports the following covariance functions (kernels):

* Exponentiated quadratic (squared exponential)
* Rational quadratic
* Periodic
* Linear
* White noise

`tsgp` flexibly enables composite kernels (either through addition or multiplication) to be constructed and is actively encouraged to appropriately model complex temporal dynamics.

`tsgp` also includes functions for computing and visualising draws from Gaussian process priors and posteriors, visualising covariance matrices, and plotting predictions.

## Performance

`tsgp` is *extremely* fast at what it does.
